---
title: "Data Dictionary — Group 1"
subtitle: "Exploring all datasets in the workshop workspace"
format:
  html:
    toc: true
    code-fold: true
    embed-resources: true
---

```{r setup, include=FALSE}
library(tidyverse)
```

This document describes every CSV file in the `data/` folder — what a single
row represents and what each column means.

---

## Group A — Yogurt Study

These files come from a study that looked at whether eating yogurt changes the
vaginal microbiome. Participants were randomly assigned to either eat yogurt
every day or stick with their usual diet. Samples were taken at two time
points: **baseline** and **after a course of antibiotics**.

### 00_sample_ids_yogurt.csv

**Each row represents:** one biological sample collected from a participant at
one time point.

```{r}
sample_ids_yog <- read_csv("../data/group_a_yogurt/00_sample_ids_yogurt.csv")
glimpse(sample_ids_yog)
```

| Column      | Meaning                                                                 |
|-------------|-------------------------------------------------------------------------|
| `pid`       | Participant ID (links to participant metadata)                          |
| `time_point`| When the sample was taken — `baseline` or `after_antibiotic`            |
| `arm`       | Which group the participant was in — `yogurt` or `unchanged_diet`       |
| `sample_id` | Unique ID for this sample (links to qPCR and Luminex results)           |

---

### 01_participant_metadata_yogurt.csv

**Each row represents:** one participant in the yogurt study.

```{r}
meta_yog <- read_csv("../data/group_a_yogurt/01_participant_metadata_yogurt.csv")
glimpse(meta_yog)
```

| Column              | Meaning                                                            |
|---------------------|--------------------------------------------------------------------|
| `pid`               | Participant ID                                                     |
| `arm`               | Study arm — `yogurt` or `unchanged_diet`                           |
| `days_since_last_sex` | Number of days since the participant last had sex                 |
| `birth_control`     | Type of birth control used — `no hormonal birth control` or `Depoprovera` |
| `age`               | Age in years                                                       |
| `education`         | Highest education level completed                                  |
| `sex`               | Sex assigned at birth — all participants are `female`              |

---

### 02_qpcr_results_yogurt.csv

**Each row represents:** a single sample's qPCR results showing how much of
three bacterial targets were detected.

```{r}
qpcr_yog <- read_csv("../data/group_a_yogurt/02_qpcr_results_yogurt.csv")
glimpse(qpcr_yog)
```

| Column            | Meaning                                                           |
|-------------------|-------------------------------------------------------------------|
| `sample_id`       | Unique sample ID (links to sample_ids table)                      |
| `qpcr_bacteria`   | Total bacterial load (copies / mL) measured by universal 16S qPCR |
| `qpcr_crispatus`  | Abundance of *Lactobacillus crispatus* (copies / mL)              |
| `qpcr_iners`      | Abundance of *Lactobacillus iners* (copies / mL)                  |

> Both *L. crispatus* and *L. iners* are species of lactobacilli that are
> common in a healthy vaginal microbiome. A high *crispatus* / low *iners*
> balance is generally considered healthier.

---

### 03_luminex_results_yogurt.csv

**Each row represents:** one cytokine measurement from one sample.

```{r}
luminex_yog <- read_csv("../data/group_a_yogurt/03_luminex_results_yogurt.csv")
glimpse(luminex_yog)
```

| Column      | Meaning                                                             |
|-------------|---------------------------------------------------------------------|
| `sample_id` | Unique sample ID (links to sample_ids table)                        |
| `cytokine`  | Name of the immune signalling protein measured (e.g. IL-1a, TNFa)   |
| `conc`      | Concentration of that cytokine in the sample                        |
| `limits`    | Whether the value was `within limits` or `out of range` (too low/high) |

The cytokines measured are: IFN-Y, IL-10, IL-1a, IL-1b, IL-6, IL-8, IP-10,
MIG, MIP-3a, and TNFa. These are **immune signalling molecules** (cytokines
and chemokines) that help regulate inflammation and immune responses.

---

## Group B — Menstruation Study

This study looked at how the vaginal microbiome and immune system change
across the menstrual cycle. Samples were taken at four time points linked to
the menstrual period.

### 00_sample_ids_period.csv

**Each row represents:** one biological sample collected from a participant at
one time point in their menstrual cycle.

```{r}
sample_ids_per <- read_csv("../data/group_b_menstruation/00_sample_ids_period.csv")
glimpse(sample_ids_per)
```

| Column      | Meaning                                                              |
|-------------|----------------------------------------------------------------------|
| `pid`       | Participant ID (links to participant metadata)                       |
| `time_point`| When in the cycle — `onset`, `end_bleeding`, `week_post`, `week_prior` |
| `arm`       | Study arm — `birth_control` or `no_birth_control`                    |
| `sample_id` | Unique sample ID (links to Luminex and flow cytometry results)       |

---

### 01_participant_metadata_period.csv

**Each row represents:** one participant in the menstruation study.

```{r}
meta_per <- read_csv("../data/group_b_menstruation/01_participant_metadata_period.csv")
glimpse(meta_per)
```

| Column           | Meaning                                                        |
|------------------|----------------------------------------------------------------|
| `pid`            | Participant ID                                                 |
| `arm`            | Study arm — `birth_control` or `no_birth_control`              |
| `age`            | Age in years                                                   |
| `pcos_status`    | Whether the participant has PCOS — `no disease` or `pcos`       |
| `period_product` | Menstrual product used — `tampon`, `pad`, or `menstrual_cup`   |
| `sex`            | Sex assigned at birth — all participants are `female`          |

> PCOS = Polycystic Ovary Syndrome, a hormonal condition that can affect
> menstrual cycles and fertility.

---

### 02_luminex_period.csv

**Each row represents:** one cytokine measurement from one sample in the
menstruation study.

```{r}
luminex_per <- read_csv("../data/group_b_menstruation/02_luminex_period.csv")
glimpse(luminex_per)
```

| Column      | Meaning                                                                   |
|-------------|---------------------------------------------------------------------------|
| `sample_id` | Unique sample ID (links to sample_ids table)                              |
| `cytokine`  | Name of the immune signalling protein measured                            |
| `conc`      | Concentration of that cytokine                                            |
| `limits`    | Whether the value was detected — `not_censored`, `below_detection_limit`, or `above_detect_limit` |

Note: the cytokines here overlap with the yogurt study but don't include
IL-8 or IL-10 (the periods dataset has fewer cytokines measured).

---

### 03_flow_cytometry_period.csv

**Each row represents:** one sample's immune cell counts measured by flow
cytometry.

```{r}
flow_per <- read_csv("../data/group_b_menstruation/03_flow_cytometry_period.csv")
glimpse(flow_per)
```

| Column              | Meaning                                                          |
|---------------------|------------------------------------------------------------------|
| `sample_id`         | Unique sample ID                                                 |
| `live_cd19_negative`| Total live cells that are CD19- (not B cells)                    |
| `cd45_negative`     | Cells that don't express CD45 (not white blood cells)            |
| `cd45_positive`     | Cells that express CD45 (white blood cells / leukocytes)         |
| `neutrophils`       | Neutrophil count (a type of innate immune cell)                  |
| `non_neutrophils`   | White blood cells that are not neutrophils                       |
| `cd3_negative`      | Cells that are CD3- (not T cells)                                |
| `cd3_positive`      | Cells that are CD3+ (T cells)                                    |
| `cd4_t_cells`       | CD4+ T cell count ("helper" T cells)                             |
| `cd8_t_cells`       | CD8+ T cell count ("killer" / cytotoxic T cells)                |

> Flow cytometry uses lasers to count and sort cells by the proteins on their
> surface (CD markers). Each CD protein marks a different type of immune cell.

---

## Instructional Dataset — UKZN Workshop 2023

This is the dataset used in the workshop. It comes from a clinical trial
testing a **treatment** vs a **placebo** for vaginal health. Samples were
taken at **baseline**, **week 1**, and **week 7**.

### 00_sample_ids_UKZN_workshop_2023.csv

**Each row represents:** one biological sample.

```{r}
sample_ids_ukzn <- read_csv("../data/instructional_dataset/00_sample_ids_UKZN_workshop_2023.csv")
glimpse(sample_ids_ukzn)
```

| Column      | Meaning                                                   |
|-------------|-----------------------------------------------------------|
| `pid`       | Participant ID                                            |
| `time_point`| When — `baseline`, `week_1`, or `week_7`                  |
| `arm`       | Study arm — `treatment` or `placebo`                      |
| `sample_id` | Unique sample ID                                          |

---

### 01_participant_metadata_UKZN_workshop_2023.csv

**Each row represents:** one participant.

```{r}
meta_ukzn <- read_csv("../data/instructional_dataset/01_participant_metadata_UKZN_workshop_2023.csv")
glimpse(meta_ukzn)
```

| Column      | Meaning                                              |
|-------------|------------------------------------------------------|
| `pid`       | Participant ID                                       |
| `arm`       | Study arm — `treatment` or `placebo`                 |
| `smoker`    | Smoking status — `smoker` or `non-smoker`            |
| `age`       | Age in years                                         |
| `education` | Highest education level completed                    |
| `sex`       | Sex — all participants are `F` (female)              |

---

### 02_visit_clinical_measurements_UKZN_workshop_2023.csv

**Each row represents:** a set of clinical measurements taken at one visit for
one participant.

```{r}
clinical_ukzn <- read_csv("../data/instructional_dataset/02_visit_clinical_measurements_UKZN_workshop_2023.csv")
glimpse(clinical_ukzn)
```

| Column        | Meaning                                                                 |
|---------------|-------------------------------------------------------------------------|
| `pid`         | Participant ID                                                          |
| `time_point`  | When — `baseline`, `week_1`, or `week_7`                                |
| `arm`         | Study arm — `treatment` or `placebo`                                    |
| `nugent_score`| Nugent score (0–10), a measure of vaginal microbiome health (lower = healthier) |
| `crp_blood`   | C-reactive protein level in blood (a marker of inflammation)            |
| `ph`          | Vaginal pH                                                              |

> The **Nugent score** is a standard clinical score for bacterial vaginosis.
> Scores of 0–3 are considered normal, 4–6 intermediate, and 7–10 indicate
> bacterial vaginosis.

---

### 03_elisa_cytokines_UKZN_workshop_2023.csv

**Each row represents:** one cytokine measurement from one sample.

```{r}
elisa_ukzn <- read_csv("../data/instructional_dataset/03_elisa_cytokines_UKZN_workshop_2023.csv")
glimpse(elisa_ukzn)
```

| Column      | Meaning                                                        |
|-------------|----------------------------------------------------------------|
| `sample_id` | Unique sample ID                                               |
| `cytokine`  | Name of the immune signalling protein measured                 |
| `conc`      | Concentration of that cytokine                                 |
| `limits`    | Whether the value was `within limits` or `out of range`        |

The cytokines here are the same 10 as in the yogurt Luminex panel: IFN-Y,
IL-10, IL-1a, IL-1b, IL-6, IL-8, IP-10, MIG, MIP-3a, and TNFa.

> ELISA (Enzyme-Linked Immunosorbent Assay) is a lab technique that measures
> the concentration of a specific protein in a sample. Luminex is a newer
> technology that can measure many proteins at once — both give similar types
> of data.

---

### 04_flow_cytometry_UKZN_workshop_2023.csv

**Each row represents:** one sample's immune cell counts measured by flow
cytometry.

```{r}
flow_ukzn <- read_csv("../data/instructional_dataset/04_flow_cytometry_UKZN_workshop_2023.csv")
glimpse(flow_ukzn)
```

| Column              | Meaning                                                           |
|---------------------|-------------------------------------------------------------------|
| `sample_id`         | Unique sample ID                                                  |
| `live_cd19_negative`| Total live cells that are CD19- (not B cells)                     |
| `cd45_negative`     | Cells that don't express CD45 (not white blood cells)             |
| `cd45_positive`     | Cells that express CD45 (white blood cells / leukocytes)          |
| `neutrophils`       | Neutrophil count                                                  |
| `non_neutrophils`   | White blood cells that are not neutrophils                        |
| `cd3_negative`      | Cells that are CD3- (not T cells)                                 |
| `cd3_positive`      | Cells that are CD3+ (T cells)                                     |
| `cd4_t_cells`       | CD4+ T helper cells                                               |
| `cd8_t_cells`       | CD8+ cytotoxic T cells                                            |

> The columns are identical to the flow cytometry data in Group B — same lab
> protocol was used across studies.

---

## Modeling Data — Sorghum Timecourse

### 01-sorghum_timecourse.csv

**Each row represents:** one sample (a batch of leaves from one plot) at one
time point, with gene expression counts for 9 genes of interest.

Note: this file is in **wide format** — the first 9 columns are gene expression
counts (with gene IDs as column names), followed by 22 columns of experimental
metadata. This is different from the other datasets which are in **long
format** (one measurement per row).

```{r}
sorghum <- read_csv("../data/modeling/01-sorghum_timecourse.csv")
glimpse(sorghum)
```

**Gene expression columns** (first 9 columns, counts from RNA sequencing):

| Column                                    | Meaning                                                |
|-------------------------------------------|--------------------------------------------------------|
| `Sobic.001G062300.v3.1` through `...`     | Expression count for each of 9 sorghum genes           |

**Experimental metadata columns:**

| Column                     | Meaning                                                          |
|----------------------------|------------------------------------------------------------------|
| `Barcode`                  | Sequencing barcode used to identify this sample                  |
| `libraryName`              | Sequencing library name (short code for the sample)              |
| `Block`                    | Field plot block number (experimental design)                    |
| `Week`                     | Week number of the experiment                                    |
| `Replicate`                | Biological replicate number (1–3)                                |
| `Genotype`                 | Sorghum variety — `RT430` or `BT642`                             |
| `Condition`                | Treatment — `Preflowering` or `Postflowering`                    |
| `Sample.type`              | Tissue type — `Leaves`, `Control`, `Preflowering`, or `Postflowering` |
| `Day`                      | Day of sampling (e.g. `Day21`)                                   |
| `Row`                      | Field row location                                               |
| `isFloweringWeek`          | Whether this was the flowering week (`TRUE` / `FALSE`)           |
| `Date.Harvested`           | Date the sample was harvested                                    |
| `Collection.Time`          | Time of day the sample was collected                             |
| `No.plants.pooled`         | How many individual plants were pooled into this sample          |
| `FloweringJulianDate`      | Julian date when the plants flowered                             |
| `FloweringDate`            | Calendar date when the plants flowered                           |
| `FloweringWeek`            | Week number of flowering                                         |
| `FloweringDay`             | Day number of flowering                                          |
| `Unreplicated`             | Whether this sample is unreplicated (`TRUE` / `FALSE`)           |
| `CompleteSampleTreatment`  | Combined treatment label (genotype + week + condition)           |
| `Timepoint`                | Numeric time point (2–14, aligns with Week)                      |
| `Group`                    | Combined group label (genotype + condition, e.g. `RT430:Preflowering`) |

> This dataset tracks **gene expression** over time in two sorghum varieties,
> looking at how gene activity changes before and after flowering. The 9 genes
> were selected because they're known to be involved in plant development or
> stress responses.
