---
title: "Exercise 1 — Yogurt Study Data"
format: html
---

## About this dataset

This file comes from a **yogurt study** that looked at whether eating yogurt
every day changes the vaginal microbiome.

Participants were randomly assigned to one of two groups:

- **yogurt** — ate yogurt daily
- **unchanged_diet** — stuck with their usual diet

Samples were taken at two time points: **baseline** (before the study started)
and **after_antibiotic** (after a course of antibiotics).

### The file: `00_sample_ids_yogurt.csv`

Each row is one biological sample. It has 4 columns:

| Column       | Meaning                                                        |
|--------------|----------------------------------------------------------------|
| `pid`        | Participant ID (links to other data files about this person)   |
| `time_point` | When the sample was taken — `baseline` or `after_antibiotic`   |
| `arm`        | Which group — `yogurt` or `unchanged_diet`                     |
| `sample_id`  | Unique ID for this sample (links to lab results)               |

Let's load it and take a look:

```{r}
library(readr)
df <- read_csv("/workspaces/Data_science_workshop/data/group_a_yogurt/00_sample_ids_yogurt.csv")
df
```

```{r}
# How many rows (samples) and columns?
rows <- nrow(df)
cols <- ncol(df)
```

The dataset has **`r rows`** samples and **`r cols`** columns:
```{r}# Load and join
library(tidyverse)
library(here)
sample_ids <- read_csv(here("data/group_a_yogurt/00_sample_ids_yogurt.csv"))
metadata <- read_csv(here("data/group_a_yogurt/01_participant_metadata_yogurt.csv"))
qpcr <- read_csv(here("data/group_a_yogurt/02_qpcr_results_yogurt.csv"))
luminex <- read_csv(here("data/group_a_yogurt/03_luminex_results_yogurt.csv"))
# Distributions and skews
#Histograms
ggplot(metadata, aes(x = age)) +
  geom_histogram(binwidth = 5, fill = "blue", color = "black") +
  labs(title = "Histogram of Age", x = "Age", y = "Frequency")

#Count NAs in each column
colSums(is.na(metadata))
#Look at the ranges with min() and max()
min(qpcr$qpcr_bacteria)
max(qpcr$qpcr_bacteria)
```
```{r}
library(gtsummary)
tbl_summary(by = group) 
```

