---
title: "Yogurt Study — Deep Dive"
format: html
editor: source
execute:
  warning: false
  message: false
---

```{r setup}
library(tidyverse)
library(here)
library(patchwork)
```

## 1. Load all four tables

Let's read every piece of the yogurt dataset and see what we're working with.

```{r load-data}
sample_ids  <- read_csv(here("data/group_a_yogurt/00_sample_ids_yogurt.csv"))
metadata    <- read_csv(here("data/group_a_yogurt/01_participant_metadata_yogurt.csv"))
qpcr        <- read_csv(here("data/group_a_yogurt/02_qpcr_results_yogurt.csv"))
luminex     <- read_csv(here("data/group_a_yogurt/03_luminex_results_yogurt.csv"))

# Quick peek at each one
glimpse(sample_ids)
```

```{r glimpse-all}
glimpse(metadata)
glimpse(qpcr)
glimpse(luminex)
```

### What one row represents in each table

| Table | One row = |
|-------|-----------|
| `sample_ids` | One biological sample (a visit by one person) |
| `metadata` | One participant (person in the study) |
| `qpcr` | One sample's bacteria counts |
| `luminex` | One sample × one cytokine measurement (10 rows per sample) |

### How they link together

```
metadata ──pid──► sample_ids ──sample_id──► qpcr
                                          ──sample_id──► luminex
```

---

## 2. Join everything into one analysis table

We'll join metadata to samples, then attach qPCR results.  
(Luminex is in *long format* — one row per cytokine — so we'll come back to it separately.)

```{r join-data}
# Join samples → metadata (via pid), then → qpcr (via sample_id)
# left_join keeps all samples even if metadata is missing
main_df <- sample_ids |>
  left_join(metadata, by = "pid") |>
  left_join(qpcr, by = "sample_id")

glimpse(main_df)
```

---

## 3. Go deep on the numbers

### 3a. Age — distribution, skew, suspicious values

```{r age-deep}
ages <- main_df |> distinct(pid, arm, age)   # one row per person

summary(ages$age)

# Histogram — quick look
ggplot(ages, aes(x = age)) +
  geom_histogram(binwidth = 2, fill = "steelblue", color = "white") +
  labs(title = "Age distribution (all participants)",
       x = "Age (years)", y = "Count")
```

**Questions to ask yourself:**

- Is the distribution symmetric or skewed? (Skewed = a long tail on one side)
- What's the youngest and oldest participant?
- Do the yogurt and control groups look similar in age?

```{r age-by-arm}
ggplot(, aes(x = arm, y = age, fill = arm)) +
  geom_boxplot() +
  labs(title = "Age by treatment arm", x = "Arm", y = "Age (years)")
```

### 3b. Days since last sex — watch for outliers

```{r sex-deep}
days <- main_df |> distinct(pid, arm, days_since_last_sex)

summary(days$days_since_last_sex)

ggplot(days, aes(x = days_since_last_sex)) +
  geom_histogram(binwidth = 2, fill = "tomato", color = "white") +
  labs(title = "Days since last sex",
       x = "Days", y = "Count")
```

**⚠️ Suspicious values:** Look at the max. Is 48 days plausible?  
Check `pid_50` in the metadata — is there anything unusual about that participant?

### 3c. Birth control — categorical counts

```{r bc-table}
main_df |>
  distinct(pid, arm, birth_control) |>
  count(arm, birth_control) |>
  pivot_wider(names_from = arm, values_from = n, values_fill = 0)
```

### 3d. Education — categorical counts

```{r edu-table}
main_df |>
  distinct(pid, arm, education) |>
  count(arm, education) |>
  pivot_wider(names_from = arm, values_from = n, values_fill = 0)
```

### 3e. qPCR bacteria counts — hugely skewed

```{r qpcr-deep}
summary(select(qpcr, qpcr_bacteria, qpcr_crispatus, qpcr_iners))
```

Those ranges are enormous — from single digits to tens of *millions*.  
That's **heavy right skew** (most samples have low values, a few have huge values).

```{r qpcr-hist}
# Total bacteria — raw scale
p1 <- ggplot(qpcr, aes(x = qpcr_bacteria)) +
  geom_histogram(bins = 30, fill = "darkgreen", color = "white") +
  labs(title = "Total bacteria (raw scale)")

# Same data on log10 scale — much easier to see
p2 <- ggplot(qpcr, aes(x = qpcr_bacteria)) +
  geom_histogram(bins = 30, fill = "darkgreen", color = "white") +
  scale_x_log10() +
  labs(title = "Total bacteria (log10 scale)")

p1 / p2
```

**Key insight:** Many biological measurements are log-normally distributed.  
The raw histogram is useless — everything piles up on the left.  
Log transform reveals the real shape.

```{r qpcr-crispatus}
# L. crispatus — note how many samples have 0
summary(qpcr$qpcr_crispatus)
sum(qpcr$qpcr_crispatus == 0)

ggplot(qpcr, aes(x = qpcr_crispatus + 0.1)) +
  geom_histogram(bins = 30, fill = "purple", color = "white") +
  scale_x_log10() +
  labs(title = "L. crispatus (log10, +0.1 to handle zeros)",
       x = "qpcr_crispatus + 0.1 (log scale)")
```

**⚠️ Suspicious:** Most samples have *zero* `qpcr_crispatus`. A few have very large numbers.  
This kind of zero-inflated distribution is common in microbiome data.

### 3f. Cytokines — data quality: "out of range" values

```{r luminex-glance}
luminex |>
  count(cytokine, limits) |>
  arrange(cytokine, limits)
```

**⚠️ Many cytokine values are "out of range"** — that means the lab instrument
couldn't measure them accurately because they were below (or above) its
detection limit. These aren't *missing*, but they're not very precise.

```{r luminex-conc}
ggplot(luminex, aes(x = conc)) +
  geom_histogram(bins = 40, fill = "orange", color = "white") +
  scale_x_log10() +
  facet_wrap(vars(cytokine), scales = "free_y") +
  labs(title = "Cytokine concentrations (log scale, all samples)")
```

---

## 4. Missing data check

```{r missing}
# How many NAs in each column of the main joined table?
main_df |>
  summarise(across(everything(), ~ sum(is.na(.x)))) |>
  pivot_longer(everything(), names_to = "column", values_to = "n_missing") |>
  filter(n_missing > 0)

# In the metadata specifically
metadata |>
  summarise(across(everything(), ~ sum(is.na(.x)))) |>
  pivot_longer(everything(), names_to = "column", values_to = "n_missing") |>
  filter(n_missing > 0)
```

**Think about this:** If there's no missing data, that's a bit suspicious in
real-world biomedical data. Why might that be? (Maybe the data was already
cleaned, or maybe missing participants were simply excluded before it reached us.)

---

## 5. Iterate on one plot three times: age by arm

The goal here is to pick one question — *"Does age look similar between the
yogurt and unchanged diet groups?"* — and try three different ways to visualize
it. You learn more by re-making one plot than by making three different plots.

### Try 1: Side-by-side histograms

```{r plot-v1}
ggplot(ages, aes(x = age, fill = arm)) +
  geom_histogram(binwidth = 3, color = "white", alpha = 0.6) +
  labs(title = "Try 1: Overlaid histograms (binwidth = 3)",
       x = "Age", y = "Count", fill = "Arm")
```

**What do you notice?** Is there good overlap? Can you tell both groups apart?

---

### Try 2: Faceted histograms + density curve

The overlaid histograms from Try 1 are hard to read because they overlap.
Let's split them into separate panels and add a smooth density curve.

```{r plot-v2}
ggplot(ages, aes(x = age, fill = arm)) +
  geom_histogram(aes(y = after_stat(density)),
                 binwidth = 2, color = "white", alpha = 0.7) +
  geom_density(alpha = 0.3) +
  facet_wrap(vars(arm), ncol = 1) +
  labs(title = "Try 2: Faceted histograms + density curves",
       x = "Age", y = "Density", fill = "Arm")
```

**What changed?** The `y = after_stat(density)` makes the y-axis a proportion
instead of raw counts — that way the two groups (which have different numbers
of people) are comparable. The density curve smooths out the bumps.

---

### Try 3: Boxplot + jittered points

```{r plot-v3}
ggplot(ages, aes(x = arm, y = age, fill = arm, color = arm)) +
  geom_boxplot(alpha = 0.4, outlier.color = NA) +
  geom_jitter(width = 0.15, size = 2, alpha = 0.7) +
  labs(title = "Try 3: Boxplot + individual points",
       x = "Arm", y = "Age (years)") +
  theme(legend.position = "none")
```

**Why this is different:**
- The boxplot shows the median (middle line), the interquartile range (box),
  and whiskers (range of typical values)
- The jittered points show *every individual person* — you can see the actual
  data, not just summaries
- You can spot outliers, clusters, and whether the groups overlap

**Which version do you prefer for telling the story of age balance between groups?**

---

## 6. Your turn: iterate more

Pick another variable and make three plots for it. For example:
- **Total bacteria (qpcr_bacteria) by arm** — raw vs log scale
- **Days since last sex by arm** — with and without that outlier
- **IL-8 concentration by time_point** — filtering out "out of range" vs keeping it

Or if you want, keep going on age — try:
- Changing binwidths in Try 1
- Adding a vertical line for the mean age with `geom_vline()`
- Coloring by `birth_control` instead of `arm`