---
title: "Regression Model — Yogurt Study"
format: html
editor: source
execute:
  warning: false
  message: false
---

> Fitting a mixed model to test the effect of yogurt on **L. iners** bacterial levels, adjusting for the imbalanced covariate `birth_control`, and accounting for repeated samples (baseline + after_antibiotic) from the same participant.

## Why this model?

| Concern | How it's handled |
|---|---|
| **Repeated measures** (2 time points per person) | Random intercept `(1 \| pid)` |
| **Imbalanced covariate** (birth_control, p=0.017 in Table 1) | Added as a fixed effect |
| **Skewed outcome** (qpcr_iners ranges 0–3.8M) | Log-transformed: `log(qpcr_iners + 1)` |

## Setup

```{r setup}
library(tidyverse)
library(here)
library(gtsummary)
library(lme4)
library(broom.mixed)
```

## Read & join data

```{r data}
samples <- read_csv(
  here("data/group_a_yogurt/00_sample_ids_yogurt.csv"),
  col_types = cols(pid = col_character())
)
meta <- read_csv(
  here("data/group_a_yogurt/01_participant_metadata_yogurt.csv"),
  col_types = cols(pid = col_character())
)
qpcr <- read_csv(
  here("data/group_a_yogurt/02_qpcr_results_yogurt.csv"),
  col_types = cols(sample_id = col_character())
)

df <- samples |>
  left_join(meta, by = c("pid", "arm")) |>
  inner_join(qpcr, by = "sample_id") |>
  mutate(
    log_iners = log(qpcr_iners + 1),       # log-transform for normality
    time_point = fct_relevel(time_point, "baseline")  # baseline as reference
  )
```

## Fit the model

We want to know: **Does the change in L. iners from baseline to after_antibiotic differ between the yogurt and unchanged-diet arms?**

```{r model}
m <- lmer(
  log_iners ~ arm * time_point + birth_control + (1 | pid),
  data = df
)
```

## Coefficient table with `tbl_regression()`

```{r table}
m |>
  tbl_regression(
    intercept = TRUE,
    label = list(
      arm           ~ "Study arm",
      time_point    ~ "Time point",
      birth_control ~ "Birth control",
      `arm:time_point` ~ "Arm × Time interaction"
    )
  ) |>
  bold_labels()
```

## Model summary (quick check)

```{r summary}
summary(m)
```