
```{r}
title: "Group A — Data Dossier"
format:
  html:
    toc: true
    code-fold: true
    embed-resources: true
```
```{r}
library(tidyverse)
library(here)
#A rendered data dossier that explains every table and shows a few plots
┌───────────────────────────────────────────┬─────────────┬────────────────────┬───────────┬─────────┐ 
 │ Characteristic                            │ Overall (N  │ unchanged_diet (N  │ yogurt (N │ p-value │ 
 │                                           │ = 51)       │ = 25)              │ = 26)     │         │ 
 ├───────────────────────────────────────────┼─────────────┼────────────────────┼───────────┼─────────┤ 
 │ days_since_last_sex                       │ 5 (3, 11)   │ 6 (3, 10)          │ 5 (3, 11) │ 0.8     │ 
 ├───────────────────────────────────────────┼─────────────┼────────────────────┼───────────┼─────────┤ 
 │ birth_control                             │             │                    │           │ 0.017   │ 
 ├───────────────────────────────────────────┼─────────────┼────────────────────┼───────────┼─────────┤ 
 │ &nbsp;&nbsp;&nbsp;&nbsp;Depoprovera       │ 25 (49%)    │ 8 (32%)            │ 17 (65%)  │         │ 
 ├───────────────────────────────────────────┼─────────────┼────────────────────┼───────────┼─────────┤ 
 │ &nbsp;&nbsp;&nbsp;&nbsp;no hormonal birth │ 26 (51%)    │ 17 (68%)           │ 9 (35%)   │         │ 
 │ control                                   │             │                    │           │         │ 
 ├───────────────────────────────────────────┼─────────────┼────────────────────┼───────────┼─────────┤ 
 │ age                                       │ 30.7 (3.3)  │ 30.4 (3.3)         │ 31.0      │ 0.5     │ 
 │                                           │             │                    │ (3.3)     │         │ 
 ├───────────────────────────────────────────┼─────────────┼────────────────────┼───────────┼─────────┤ 
 │ education                                 │             │                    │           │ 0.8     │ 
 ├───────────────────────────────────────────┼─────────────┼────────────────────┼───────────┼─────────┤ 
 │ &nbsp;&nbsp;&nbsp;&nbsp;grade 10–12,      │ 10 (20%)    │ 6 (24%)            │ 4 (15%)   │         │ 
 │ matriculated                              │             │                    │           │         │ 
 ├───────────────────────────────────────────┼─────────────┼────────────────────┼───────────┼─────────┤ 
 │ &nbsp;&nbsp;&nbsp;&nbsp;grade 10–12, not  │ 21 (41%)    │ 9 (36%)            │ 12 (46%)  │         │ 
 │ matriculated                              │             │                    │           │         │ 
 ├───────────────────────────────────────────┼─────────────┼────────────────────┼───────────┼─────────┤ 
 │ &nbsp;&nbsp;&nbsp;&nbsp;less than grade 9 │ 9 (18%)     │ 4 (16%)            │ 5 (19%)   │         │ 
 ├───────────────────────────────────────────┼─────────────┼────────────────────┼───────────┼─────────┤ 
 │ &nbsp;&nbsp;&nbsp;&nbsp;post-secondary    │ 11 (22%)    │ 6 (24%)            │ 5 (19%)   │         │ 
 └───────────────────────────────────────────┴─────────────┴────────────────────┴───────────┴─────────┘ 

