---
title: "Table 1 — Yogurt Study Participant Demographics"
format: html
editor: source
execute:
  warning: false
  message: false
---

> Creating a **Table 1** — a summary table of participant characteristics, split by study arm (yogurt vs unchanged diet), with p-values from `gtsummary`.

## Setup

Load the packages and read in the data:

```{r setup}
library(tidyverse)
library(here)
library(gtsummary)

df <- read_csv(
  here("data/group_a_yogurt/01_participant_metadata_yogurt.csv"),
  col_types = cols(pid = col_character())
)
```

## Quick peek at the data

```{r}
df
```

## Step 1 — Basic `tbl_summary` by arm

Remove the participant ID column (`pid`) since it's just an identifier, then summarise everything else split by the `arm` column (yogurt vs unchanged_diet).

```{r step1}
df |>
  select(-pid) |>
  tbl_summary(by = arm)
```

**What it shows:** For each variable, it shows the distribution within each arm — counts (n, %) for categorical variables (birth_control, education, sex) and median (Q1, Q3) for continuous variables (days_since_last_sex, age).

---

## Step 2 — Add p-values with `add_p()`

This adds a p-value column testing whether each variable differs between the two arms. `gtsummary` picks the right test automatically (e.g. Wilcoxon rank-sum for continuous, chi-square for categorical).

```{r step2}
df |>
  select(-pid) |>
  tbl_summary(by = arm) |>
  add_p()
```

---

## Step 3 — Override `age` to show mean (SD) and use a t-test

By default `tbl_summary` shows median (Q1, Q3) for continuous variables. Let's change `age` to mean (SD) and use a t-test instead of the Wilcoxon test.

```{r step3}
df |>
  select(-pid) |>
  tbl_summary(
    by = arm,
    statistic = list(age ~ "{mean} ({sd})")
  ) |>
  add_p(test = list(age ~ "t.test"))
```

---

## Step 4 — Polish: add overall column, bold labels, nicer header

Add the finishing touches: an overall column that shows the whole-sample distribution, bold variable labels, and a cleaner header.

```{r step4}
df |>
  select(-pid) |>
  tbl_summary(
    by = arm,
    statistic = list(age ~ "{mean} ({sd})")
  ) |>
  add_p(test = list(age ~ "t.test")) |>
  add_overall() |>
  bold_labels() |>
  modify_header(label ~ "**Characteristic**")
```

---

## Quick interpretation

- **Which variables look balanced** between the yogurt and unchanged-diet groups? (Look for large p-values.)
- **Which variables might be imbalanced?** (Look for small p-values.) Could any of these confound the yogurt effect?