---
title: "Table 1 — Yogurt Study (Manual Overrides)"
format: html
editor: source
execute:
  warning: false
  message: false
---

> Creating a Table 1 with `gtsummary`, going **variable by variable** to decide whether the default summary and default test are appropriate, and overriding where they aren't.

## Variables & decisions

| Variable | Type | Default | Decision |
|---|---|---|---|
| `pid` | ID | — | **Exclude** |
| `arm` | Group | — | **`by` variable** |
| `days_since_last_sex` | Count (days) | Median (IQR), Wilcoxon | ✅ Keep — skewed count data |
| `birth_control` | Nominal | n (%), chi-square | ✅ Keep |
| `age` | Continuous (roughly normal) | Median (IQR), Wilcoxon | **Override** → mean (SD), t-test |
| `education` | Ordinal (4 levels) | n (%), chi-square | ✅ Keep — reasonable for Table 1 |
| `sex` | Constant (all female) | — | **Drop** — no variation |

## Setup

```{r setup}
library(tidyverse)
library(here)
library(gtsummary)

df <- read_csv(
  here("data/group_a_yogurt/01_participant_metadata_yogurt.csv"),
  col_types = cols(pid = col_character())
)
```

## Build the table with overrides

```{r table}
df |>
  select(-pid, -sex) |>                        # drop ID + constant column
  tbl_summary(
    by = arm,
    statistic = list(age ~ "{mean} ({sd})")     # age: mean (SD) instead of median (IQR)
  ) |>
  add_p(
    test = list(age ~ "t.test")                 # age: t-test instead of Wilcoxon
  ) |>
  add_overall() |>                              # add an "Overall" column
  bold_labels()                                 # bold variable names
```

## Quick check

```{r quick-check}
# Verify mean & SD for age in each arm
df |>
  group_by(arm) |>
  summarise(
    mean_age = mean(age),
    sd_age   = sd(age),
    n        = n()
  )
```
